Scalable Clustering of Genotype Information using MapReduce

### 2014-05-17 doc/publications/incob2014/paper.tex
addition

% Our variable names
\newcommand{\chrOnePop}{Chr1\_Population}
\newcommand{\FullPop}{Full\_Population}
\newcommand{\FullFam}{Chr1\_Family}

@@

\subsection*{Accuracy}
Text for this section.

@@

\subsection*{R K-means clustering}
We use 'readVcf' from the R library, 'VariantAnnotation', to iteratively read in the VCF file. For each iteration, we convert the data to a matrix and
change the values for each variant to the Hamming distance, as mentioned previously. We then transpose the matrix, so individuals are represented
by rows, and append the matrix to that from the previous iteration. Once the matrix contains all of the variants from the VCF file, we cluster the data
using the 'kmeans' algorithm from the 'stats' library.

\subsection*{Clustering quality}
We scored the clusters with the Adjusted Rand index. 
%TODO: can you provide the formular
This index compares two different clusterings and assigns a value based on their similarity. The value is zero for independent clusterings and one for identical clusterings. We compared the Mahout k-means
clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project. For the known clusters,
individuals were partitioned by their super population code (ASN, EUR, AFR, AMR). 
%TODO: "also on \chrOnePop" on what dataset did you compare the R performance otherwise ?
We also compared the Mahout clusters to the
R clusters for the Chr1 subset data.

\subsection*{Speed}
We measured the speed for various stages of the processing and clustering. We used

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Datasets used in this study.}
      \begin{tabular}{lcc}
        \hline
           & individuals  & variants  \\ \hline
        \chrOnePop & 1024 & 3,007,196 \\
        \FullPop\ & 1024 & 38,219,238 \\
        \FullFam\ & 1024 & ...\\
      \end{tabular}
      \label{datasets}
\end{table}

\begin{table}[h!]
\caption{Adjusted Rand index (accuracy).}
      \begin{tabular}{lccc}
        \hline
           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
        Mahout, k=4 & 0.664 & ..&..\\
        Mahout, k=10 & 0.714 & ..&..\\
        Mahout, k=20 & 0.769 & .. & ..\\ 
        Mahout, k=30 & 0.728 & .. & ..\\ 
        R, k=4 & .. & - & -\\
        R, k=10 & .. & - & -\\
        R, k=20 & .. & - & -\\ 
        R, k=30 & .. & - & -\\ \hline
      \end{tabular}
      \label{datasetsAcc}
\end{table}

\begin{table}[h!]
\caption{Time taken to cluster samples (k=20).}
      \begin{tabular}{lccc}
        \hline
           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
        Mahout, 1 node & .. & ..&..\\
        Mahout, 2 nodes & .. & ..&..\\
        Mahout, 3 nodes & .. & .. & ..\\ 
        Mahout, 4 nodes & .. & .. & ..\\ 
        R & .. & - & -\\ \hline
      \end{tabular}
      \label{datasetsTime}
\end{table}

>>>>

% Our variable names
\newcommand{\chrOnePop}{Chr19\_Population}
\newcommand{\FullPop}{Full\_Population}
\newcommand{\FullFam}{Chr1\_Family}

@@

\subsection*{Accuracy}
For this example, we knew the correct clusters from 1000 Genomes metadata. This metadata assigns a population and family value to each of the 1092 individuals.
With the foreknowledge of the correct clusters, the Adjusted Rand Index enables us to assign a score to cluster accuracy~\cite{Hubert1985}.

@@

\subsection*{R K-means clustering}
We use `readVcf' from the R library, `VariantAnnotation', to iteratively read in the VCF file. For each iteration, we convert the data to a matrix and
change the values for each variant to the Hamming distance, as mentioned previously. We then transpose the matrix, so individuals are represented
by rows, and append the matrix to that from the previous iteration. Once the matrix contains all of the variants from the VCF file, we cluster the data
using the `kmeans' algorithm from the `stats' library.

\subsection*{Clustering quality}
We scored the clusters with the Adjusted Rand index. 
%TODO: can you provide the formular
This index compares two different cluster sets and assigns a value based on their similarity. The value is zero for independent clusterings and one for identical clusterings. We compared the Mahout k-means
clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project. For the known clusters,
individuals were partitioned by their super population code (ASN, EUR, AFR, AMR). 
%TODO: "also on \chrOnePop" on what dataset did you compare the R performance otherwise ?
We also compared the Mahout clusters to the
R clusters for the Chr1 subset data.

\subsection*{Performance}
We recorded the memory usage and run-time from the PBS logs. For the Mahout jobs, we queued separate jobs
for pre-processing and k-means clustering to get an accurate representation of the time and memory requirements for each stage.
For the R job, where we could not easily separate the pre-processing and clustering components, we printed the time between
stages using the function, \texttt{proc.time()}.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Datasets used in this study.}
      \begin{tabular}{lcc}
        \hline
           & individuals  & variants  \\ \hline
        \chrOnePop & 1092 & 816,115 \\
        \FullPop\ & 1092 & 38,219,238 \\
        \FullFam\ & 1092 & ...\\
      \end{tabular}
      \label{datasets}
\end{table}

\begin{table}[h!]
\caption{Adjusted Rand index (accuracy).}
      \begin{tabular}{lcccc}
        \hline
           & \chrOnePop & \FullPop  & hapmap \\ \hline
        Mahout, k=4 & 0.664 & ..&..\\
        Mahout, k=10 & 0.714 & ..&..\\
        Mahout, k=20 & 0.769 & .. & ..\\ 
        Mahout, k=30 & 0.728 & .. & ..\\ 
        R, k=4 & .. & - & -\\
        R, k=10 & .. & - & -\\
        R, k=20 & .. & - & -\\ 
        R, k=30 & .. & - & -\\ \hline
      \end{tabular}
      \label{datasetsAcc}
\end{table}

\begin{table}[h!]
\caption{Resources consumed.}
      \begin{tabular}{lccccc}
      \hline
      &&\multicolumn{2}{ c }{Memory (kb)}&\multicolumn{2}{ c }{Time (s)} \\
        \hline
           & Nodes & pmem & vmem  & pre-processing & clustering \\ \hline
        R, \chrOnePop & 1 & 17696940 & 18000196 &3341 & 7\\
        Mahout,\chrOnePop & 1 & 1831160 & 7646232 & 1255 & 594\\
        Mahout, \FullPop & .. &.. & .. & ..&..\\ 
        Mahout, 4 nodes & .. & .. & .. & ..&..\\ 
        \hline
      \end{tabular}
      \label{datasetsTime}
\end{table}
